---
title: "Immunization"
subtitle: "National Immunization Program Information System (SI-PNI)"
format:
  html:
    page-layout: full
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show code"
lang: en
---

```{r}
#| label: setup
#| include: false

library(crosstalk)
library(plotly)
library(reactable)
library(readr)
library(dplyr, warn.conflicts = FALSE)

source("_viz_utils.R")

# read pre-processed data
dados_brasil <- tryCatch(
  read_csv("data/imunizacao/brasil.csv", show_col_types = FALSE),
  error = function(e) tibble()
)
dados_uf <- tryCatch(
  read_csv("data/imunizacao/uf.csv", show_col_types = FALSE),
  error = function(e) tibble()
)
dados_municipal <- tryCatch(
  read_csv("data/imunizacao/municipal.csv", show_col_types = FALSE),
  error = function(e) tibble()
)

ultimo_ano <- if (nrow(dados_brasil) > 0) max(dados_brasil$ano, na.rm = TRUE) else NA
has_data <- nrow(dados_brasil) > 0
```

Vaccination coverage data for the main immunobiologicals in Brazil's national vaccination schedule, based on SI-PNI (DATASUS). Indicators include vaccination coverage by immunobiological, state and municipality comparisons.

**Source:** SI-PNI/DATASUS | **Period:** 2010-2024 | **Updates:** subject to data availability

::: {.panel-tabset}

## Brazil

```{r}
#| label: setup-brasil
#| include: false

br <- dados_brasil |> filter(desagregacao == "imunobiologico")
```

::: {.panel-tabset}

### Overview

```{r}
#| label: brasil-overview

if (has_data && nrow(br) > 0) {
  br_wide <- br |>
    select(ano, categoria, taxa) |>
    filter(!is.na(taxa))

  if (nrow(br_wide) > 0) {
    plot_ly(
      br_wide,
      x = ~ano,
      y = ~categoria,
      z = ~taxa,
      type = "heatmap",
      colorscale = list(
        list(0, "#DBEAFE"),
        list(0.5, "#3B82F6"),
        list(1, "#1E3A8A")
      ),
      hovertemplate = paste(
        "<b>%{y}</b><br>",
        "Year: %{x}<br>",
        "Coverage: %{z:.1f}%",
        "<extra></extra>"
      )
    ) |>
      layout_padrao(
        titulo = "Vaccination coverage by immunobiological and year",
        eixo_x = "Year",
        eixo_y = "Immunobiological"
      )
  } else {
    plot_ly() |> layout(title = "Data not available")
  }
} else {
  plot_ly() |> layout(title = "Run prep_imunizacao.R to generate data")
}
```

```{r}
#| label: brasil-table-overview

if (has_data && nrow(br) > 0) {
  tab_br <- br |>
    filter(!is.na(taxa)) |>
    select(Year = ano, Immunobiological = categoria, `Coverage (%)` = taxa) |>
    mutate(`Coverage (%)` = round(`Coverage (%)`, 1)) |>
    arrange(desc(Year), Immunobiological)

  reactable(
    tab_br,
    searchable = TRUE,
    striped = TRUE,
    highlight = TRUE,
    defaultPageSize = 10,
    columns = list(
      `Coverage (%)` = colDef(
        style = function(value) {
          if (is.na(value)) return()
          color <- if (value >= 95) "#16A34A" else if (value >= 80) "#EAB308" else "#DC2626"
          list(color = color, fontWeight = "bold")
        }
      )
    )
  )
}
```

### By Immunobiological

```{r}
#| label: brasil-by-imuno

if (has_data && nrow(br) > 0) {
  sd_br_imuno <- SharedData$new(
    br |> filter(!is.na(taxa)),
    key = ~paste(ano, categoria)
  )

  filter_select("filtro_imuno_br_en", "Select immunobiological:", sd_br_imuno, ~categoria)

  plot_ly(
    sd_br_imuno,
    x = ~ano,
    y = ~taxa,
    color = ~categoria,
    colors = cores_imuno,
    type = "scatter",
    mode = "lines+markers",
    hovertemplate = paste(
      "<b>%{fullData.name}</b><br>",
      "Year: %{x}<br>",
      "Coverage: %{y:.1f}%",
      "<extra></extra>"
    )
  ) |>
    layout_padrao(
      titulo = "Vaccination coverage time series",
      eixo_x = "Year",
      eixo_y = "Coverage (%)"
    ) |>
    layout(
      shapes = list(
        list(
          type = "line", x0 = 0, x1 = 1,
          y0 = 95, y1 = 95,
          xref = "paper", yref = "y",
          line = list(color = "#DC2626", dash = "dash", width = 1)
        )
      ),
      annotations = list(
        list(
          x = 1, y = 95, xref = "paper", yref = "y",
          text = "Target 95%", showarrow = FALSE,
          font = list(color = "#DC2626", size = 10),
          xanchor = "right"
        )
      )
    )

  reactable(
    sd_br_imuno,
    columns = list(
      ano = colDef(name = "Year"),
      categoria = colDef(name = "Immunobiological"),
      taxa = colDef(name = "Coverage (%)", format = colFormat(digits = 1)),
      periodo = colDef(show = FALSE),
      nivel_geo = colDef(show = FALSE),
      cod_local = colDef(show = FALSE),
      nome_local = colDef(show = FALSE),
      uf = colDef(show = FALSE),
      regiao = colDef(show = FALSE),
      populacao = colDef(show = FALSE),
      desagregacao = colDef(show = FALSE),
      numerador = colDef(show = FALSE),
      denominador = colDef(show = FALSE),
      unidade = colDef(show = FALSE)
    ),
    searchable = TRUE,
    striped = TRUE,
    highlight = TRUE,
    defaultPageSize = 10
  )
} else {
  cat("Run `source('data-prep/prep_imunizacao.R')` to generate data.")
}
```

:::

## By State

```{r}
#| label: setup-uf
#| include: false

uf <- dados_uf |> filter(desagregacao == "imunobiologico")
has_uf <- nrow(uf) > 0
```

::: {.panel-tabset}

### Overview

```{r}
#| label: uf-overview

if (has_uf) {
  uf_ultimo <- uf |>
    filter(ano == max(ano, na.rm = TRUE), !is.na(taxa))

  sd_uf_geral <- SharedData$new(
    uf_ultimo,
    key = ~paste(uf, categoria)
  )

  filter_select("filtro_imuno_uf_g_en", "Select immunobiological:", sd_uf_geral, ~categoria)

  plot_ly(
    sd_uf_geral,
    y = ~reorder(uf, taxa),
    x = ~taxa,
    color = ~categoria,
    colors = cores_imuno,
    type = "bar",
    orientation = "h",
    hovertemplate = paste(
      "<b>%{y}</b><br>",
      "Coverage: %{x:.1f}%",
      "<extra></extra>"
    )
  ) |>
    layout_padrao(
      titulo = paste("Vaccination coverage by state -", max(uf$ano, na.rm = TRUE)),
      eixo_x = "Coverage (%)",
      eixo_y = ""
    ) |>
    layout(
      shapes = list(
        list(
          type = "line", x0 = 95, x1 = 95,
          y0 = 0, y1 = 1,
          xref = "x", yref = "paper",
          line = list(color = "#DC2626", dash = "dash", width = 1)
        )
      ),
      margin = list(l = 50)
    )

  reactable(
    sd_uf_geral,
    columns = list(
      uf = colDef(name = "State"),
      categoria = colDef(name = "Immunobiological"),
      taxa = colDef(name = "Coverage (%)", format = colFormat(digits = 1)),
      regiao = colDef(name = "Region"),
      ano = colDef(show = FALSE),
      periodo = colDef(show = FALSE),
      nivel_geo = colDef(show = FALSE),
      cod_local = colDef(show = FALSE),
      nome_local = colDef(show = FALSE),
      populacao = colDef(show = FALSE),
      desagregacao = colDef(show = FALSE),
      numerador = colDef(show = FALSE),
      denominador = colDef(show = FALSE),
      unidade = colDef(show = FALSE)
    ),
    searchable = TRUE,
    striped = TRUE,
    highlight = TRUE,
    defaultPageSize = 27
  )
} else {
  cat("State-level data not available.")
}
```

### By Immunobiological

```{r}
#| label: uf-by-imuno

if (has_uf) {
  sd_uf_imuno <- SharedData$new(
    uf |> filter(!is.na(taxa)),
    key = ~paste(ano, uf, categoria)
  )

  bscols(
    widths = c(6, 6),
    filter_select("filtro_uf_imuno_en", "Select state:", sd_uf_imuno, ~uf),
    filter_select("filtro_imuno_uf_en", "Select immunobiological:", sd_uf_imuno, ~categoria)
  )

  plot_ly(
    sd_uf_imuno,
    x = ~ano,
    y = ~taxa,
    color = ~categoria,
    colors = cores_imuno,
    type = "scatter",
    mode = "lines+markers",
    hovertemplate = paste(
      "<b>%{fullData.name}</b><br>",
      "Year: %{x}<br>",
      "Coverage: %{y:.1f}%",
      "<extra></extra>"
    )
  ) |>
    layout_padrao(
      titulo = "Vaccination coverage time series by state",
      eixo_x = "Year",
      eixo_y = "Coverage (%)"
    ) |>
    layout(
      shapes = list(
        list(
          type = "line", x0 = 0, x1 = 1,
          y0 = 95, y1 = 95,
          xref = "paper", yref = "y",
          line = list(color = "#DC2626", dash = "dash", width = 1)
        )
      )
    )

  reactable(
    sd_uf_imuno,
    columns = list(
      ano = colDef(name = "Year"),
      uf = colDef(name = "State"),
      categoria = colDef(name = "Immunobiological"),
      taxa = colDef(name = "Coverage (%)", format = colFormat(digits = 1)),
      regiao = colDef(name = "Region"),
      periodo = colDef(show = FALSE),
      nivel_geo = colDef(show = FALSE),
      cod_local = colDef(show = FALSE),
      nome_local = colDef(show = FALSE),
      populacao = colDef(show = FALSE),
      desagregacao = colDef(show = FALSE),
      numerador = colDef(show = FALSE),
      denominador = colDef(show = FALSE),
      unidade = colDef(show = FALSE)
    ),
    searchable = TRUE,
    striped = TRUE,
    highlight = TRUE,
    defaultPageSize = 15
  )
} else {
  cat("State-level data not available.")
}
```

:::

## By Municipality

```{r}
#| label: setup-municipal
#| include: false

mun <- dados_municipal |> filter(desagregacao == "imunobiologico")
has_mun <- nrow(mun) > 0
```

::: {.panel-tabset}

### Overview

```{r}
#| label: mun-overview

if (has_mun) {
  ultimo_periodo <- max(mun$periodo, na.rm = TRUE)

  mun_ultimo <- mun |>
    filter(periodo == ultimo_periodo, !is.na(taxa))

  sd_mun <- SharedData$new(
    mun_ultimo,
    key = ~paste(cod_local, categoria)
  )

  bscols(
    widths = c(4, 4, 4),
    filter_select("filtro_uf_mun_en", "Select state:", sd_mun, ~uf),
    filter_select("filtro_imuno_mun_en", "Select immunobiological:", sd_mun, ~categoria),
    filter_slider("filtro_pop_mun_en", "Minimum population:", sd_mun, ~populacao, width = "100%")
  )

  plot_ly(
    sd_mun,
    y = ~reorder(nome_local, taxa),
    x = ~taxa,
    type = "bar",
    orientation = "h",
    marker = list(color = cores_dashboard$primaria),
    hovertemplate = paste(
      "<b>%{y}</b><br>",
      "Coverage: %{x:.1f}%",
      "<extra></extra>"
    )
  ) |>
    layout_padrao(
      titulo = paste("Vaccination coverage by municipality -", ultimo_periodo),
      eixo_x = "Coverage (%)",
      eixo_y = ""
    ) |>
    layout(
      shapes = list(
        list(
          type = "line", x0 = 95, x1 = 95,
          y0 = 0, y1 = 1,
          xref = "x", yref = "paper",
          line = list(color = "#DC2626", dash = "dash", width = 1)
        )
      ),
      margin = list(l = 120)
    )

  reactable(
    sd_mun,
    columns = list(
      cod_local = colDef(name = "IBGE Code"),
      nome_local = colDef(name = "Municipality"),
      uf = colDef(name = "State"),
      categoria = colDef(name = "Immunobiological"),
      taxa = colDef(name = "Coverage (%)", format = colFormat(digits = 1)),
      populacao = colDef(name = "Population"),
      periodo = colDef(name = "Period"),
      ano = colDef(show = FALSE),
      nivel_geo = colDef(show = FALSE),
      regiao = colDef(show = FALSE),
      desagregacao = colDef(show = FALSE),
      numerador = colDef(show = FALSE),
      denominador = colDef(show = FALSE),
      unidade = colDef(show = FALSE)
    ),
    searchable = TRUE,
    striped = TRUE,
    highlight = TRUE,
    defaultPageSize = 20
  )
} else {
  cat("Municipal data not available. Select a state to filter.")
}
```

:::

:::

---

### Technical notes

**Indicators:**

- **Vaccination coverage (IMU-01):** proportion of the target population that received the complete vaccination schedule for each immunobiological, calculated as (doses applied / target population) x 100.
- **PNI target:** 95% for most immunobiologicals; 90% for BCG; 80% for HPV.

**Limitations:**

- Data from 2020 onwards may present inconsistencies due to the SI-PNI system migration to the new e-SUS system.
- Vaccination coverage may exceed 100% in small municipalities due to vaccination of residents from other municipalities.
- Municipal data is presented as three-year averages for greater statistical stability.

**Source:** SI-PNI/DATASUS, processed with the [healthbR](https://github.com/SidneyBissoli/healthbR) package.
